Instant Continuation Marks--Just Add $\lambda$

### 2012-10-02 paper.tex
About to change to direct style.

In order to formalize continuation marks, we introduce an extension of the \lc with facilities to manipulate continuation marks and present its semantics in the style of Felleisen and Hieb \cite{felleisen1992revised}. Because our ultimate goal is a desugaring transform, we first introduce the target language of the transform, the \lc, in the same way.

@@

\setkeyword{cons snd}
\setspecialsymbol{lambda}{$\lambda$}
\setspecialsymbol{e_0}{$e_0$}
\setspecialsymbol{e_0p}{$e'_0$}
\setspecialsymbol{Ce_0}{$\C{e_0}$}
\setspecialsymbol{Ce_0p}{$\C{e'_0}$}
\setspecialsymbol{x_1}{$x_1$}
\setspecialsymbol{x_n}{$x_n$}
\setspecialsymbol{e_1}{$e_1$}
\setspecialsymbol{e_n}{$e_n$}
\setspecialsymbol{Ce_1}{$\C{e_1}$}
\setspecialsymbol{Ce_1p}{$\C{e'_1}$}
\setspecialsymbol{Cv}{$\C{v}$}
\setspecialsymbol{v_0}{$v_0$}
\setspecialsymbol{Cv_0}{$\C{v_0}$}
\setspecialsymbol{Cv_0p}{$\C{v'_0}$}
\setspecialsymbol{Cpv_0}{$\Cp{v_0}$}
\setspecialsymbol{Cpv_1}{$\Cp{v_1}$}
\setspecialsymbol{v_1}{$v_1$}
\setspecialsymbol{T}{$\mathcal{A}$}
\setspecialsymbol{C}{$\mathcal{C}$}
\setspecialsymbol{Cp}{$\mathcal{C}'$}
\setspecialsymbol{hole}{$\hole$}
\setspecialsymbol{bottom}{$\perp$}
\setspecialsymbol{Ce_0p_x_Cpv_1}{$\C{e'_0}[x\leftarrow\Cp{v_1}]$}
\setspecialsymbol{e_0p_x_v_1}{$e'_0[x\leftarrow v_1]$}
\setspecialsymbol{Cz}{$\C{z}$}
\setspecialsymbol{CE}{$\C{E}$}
\setspecialsymbol{xiE}{$\xi(E)$}
\setspecialsymbol{chiE}{$\chi(E)$}
\setspecialsymbol{CchiE}{$\C{\chi(E)}$}

@@

With this in mind, consider a conceptual transformation \scheme|C[(rator-expr rand-expr)]| as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    (let ((rator-value ((C[rator-expr] FALSE) marks))
          (rand-value ((C[rand-expr] FALSE) marks))
      (((rator-value rand-value) flag) marks))))
\end{schemedisplay}
ignoring for the moment that \scheme'let' is in neither \lv\ or \cm. This definition captures that
\begin{enumerate}
\item before evaluation, we expect \scheme'flag' to indicate tail position information and \scheme'marks' to provide a list of the current continuation marks,
\item we would like to evaluate \scheme'C[rator-expr]' and \scheme'C[rand-expr]' in the same manner, providing to each its contextual information--specifically that neither is evaluated in tail position of a \scheme'wcm' directive and the continuation marks for each are unchanged from the parent context,
\item and, following evaluation of operator and operand and application, evaluation of the resultant term is performed with the original contextual information.
\end{enumerate}

Now consider a conceptual transformation \scheme|C[(wcm mark-expr body-expr)]| as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    ((C[body-expr] TRUE) (let ((mark-value ((C[mark-expr] FALSE) marks))
                               (rest-marks (if flag (snd marks) marks)))
                           (cons mark-value rest-marks)))))
\end{schemedisplay}
with similar caveats as the previous case. This definition captures that
\begin{enumerate}
\item as in application, we expect \scheme'flag' to indicate tail position information and \scheme'marks' to provide a list of the current continuation marks,
\item we evaluate \scheme'mark-expr' with correct contextual information,
\item we discard the first continuation mark of the parent context if evaluation is occurring in tail position of a \scheme'wcm' directive,
\item and we evaluate \scheme'C[body-expr]' with the correct tail-position flag and current continuation marks.
\end{enumerate}

Finally, consider the conceptual transformation \scheme|C[(ccm)]| as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    marks))
\end{schemedisplay}
wherein we reap the fruits of simplicity from our laborious passing: this definition is gratifyingly direct.

@@

We can express the \scheme'let' construct in \lv\ with application. An expression such as
\begin{schemedisplay}
(let ((x_1 e_1)
      ...
      (x_n e_n))
  e)
\end{schemedisplay}
can be interpreted as
\begin{schemedisplay}
(...((lambda (x_1) ... (lambda (x_n)
           e)...) e_1) ...) e_n)
\end{schemedisplay}
which is the curried form of
\begin{schemedisplay}
((lambda (x_1 ... x_n)
           e) e_1 ... e_n)
\end{schemedisplay}

@@

Abstracting terms has the effect of suspending evaluation. When an entire program is transformed, all evaluation is suspended. It can be initiated by providing contextual information. At the top level, the context is empty, so we pass the contextual information for the empty context: \scheme'FALSE' indicating evaluation is \emph{not} occurring in \scheme'wcm' tail position and \scheme'NIL', an empty list of marks.

@@

For instance, consider the simple program \scheme|(ccm)|. $\Cp{\evalcm{\ccmb}}=\Cp{\nil}$, but
\begin{align*}
  &\evalv{\app{\app{\C{\ccmb}}{\false}}{\nil}}\\
= &\evalv{\app{\app{\abs{\mathit{flag}}{\abs{\mathit{marks}}{\mathit{marks}}}}{\false}}{\nil}}\\
= &\nil
\end{align*}
This is a problem because $\nil$ is not a primitive value but shorthand for the Church encoding of \scheme'nil', $\abs{x}{\abs{y}{y}}$. As an abstraction, it too is transformed. It is clear then that we must alter the commutativity requirement, as well as the transform, once more. The final commutativity property we seek is
\begin{equation}
\app{\app{\C{\evalcm{p}}}{\false}}{\C{\nil}}=\evalv{\app{\app{\C{p}}{\false}}{\C{\nil}}}
\end{equation}

@@

\subsection{Useful Shorthand}

We first define a useful shorthand.

@@

\begin{schemedefn}{\scheme|C[(wcm mark-expr body-expr)]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (((C[mark-expr]
         (lambda (mark-value) ((lambda (k) 
                                 (k (((marks (lambda (x) x)) (lambda (z) z)) (lambda (z) z))))
                               (lambda (t) 
                                 ((lambda (k) (k ((f C[(SND t)]) t)))
                                   (lambda (rest) 
                                     (((C[body-expr] kont) TRUE) C[(CONS mark-value rest)])))))))
        FALSE)
       marks))))
\end{schemedisplay}
\end{schemedefn}

>>>>

In order to formalize continuation marks, we introduce an extension of the \lc\ with facilities to manipulate continuation marks and present its semantics in the style of Felleisen and Hieb \cite{felleisen1992revised}. Because our ultimate goal is a desugaring transform, we first introduce the target language of the transform, the \lc, in the same way.

@@

\setkeyword{cons snd}
\setspecialsymbol{lambda}{$\lambda$}
\setspecialsymbol{e_0}{$e_0$}
\setspecialsymbol{e_0p}{$e'_0$}
\setspecialsymbol{Ce_0}{$\C{e_0}$}
\setspecialsymbol{Ce_0p}{$\C{e'_0}$}
\setspecialsymbol{x_1}{$x_1$}
\setspecialsymbol{x_n}{$x_n$}
\setspecialsymbol{e_1}{$e_1$}
\setspecialsymbol{e_n}{$e_n$}
\setspecialsymbol{k_1}{$k_1$}
\setspecialsymbol{k_2}{$k_2$}
\setspecialsymbol{Ce_1}{$\C{e_1}$}
\setspecialsymbol{Ce_1p}{$\C{e'_1}$}
\setspecialsymbol{Cv}{$\C{v}$}
\setspecialsymbol{v_0}{$v_0$}
\setspecialsymbol{Cv_0}{$\C{v_0}$}
\setspecialsymbol{Cv_0p}{$\C{v'_0}$}
\setspecialsymbol{Cpv_0}{$\Cp{v_0}$}
\setspecialsymbol{Cpv_1}{$\Cp{v_1}$}
\setspecialsymbol{v_1}{$v_1$}
\setspecialsymbol{T}{$\mathcal{A}$}
\setspecialsymbol{C}{$\mathcal{C}$}
\setspecialsymbol{Cp}{$\mathcal{C}'$}
\setspecialsymbol{hole}{$\hole$}
\setspecialsymbol{bottom}{$\perp$}
\setspecialsymbol{Ce_0p_x_Cpv_1}{$\C{e'_0}[x\leftarrow\Cp{v_1}]$}
\setspecialsymbol{e_0p_x_v_1}{$e'_0[x\leftarrow v_1]$}
\setspecialsymbol{Cz}{$\C{z}$}
\setspecialsymbol{CE}{$\C{E}$}
\setspecialsymbol{xiE}{$\xi(E)$}
\setspecialsymbol{chiE}{$\chi(E)$}
\setspecialsymbol{CchiE}{$\C{\chi(E)}$}

@@

\subsection{Concept}

The definition of $\mathcal{C}$ entails a defined transformation of each syntactic form of \cm.

With this in mind, consider a conceptual transformation of application, \scheme|C[(rator-expr rand-expr)]|, as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    (let ((rator-value ((C[rator-expr] FALSE) marks))
          (rand-value ((C[rand-expr] FALSE) marks))
      (((rator-value rand-value) flag) marks))))
\end{schemedisplay}
ignoring for the moment that \scheme'let' is in neither \lv\ or \cm. This definition captures that
\begin{enumerate}
\item before evaluation, we expect \scheme'flag' to indicate tail position information and \scheme'marks' to provide a list of the current continuation marks,
\item we would like to evaluate \scheme'C[rator-expr]' and \scheme'C[rand-expr]' in the same manner, providing to each its contextual information--specifically that neither is evaluated in tail position of a \scheme'wcm' directive and the continuation marks for each are unchanged from the parent context,
\item and, following evaluation of operator and operand and application, evaluation of the resultant term is performed with the original contextual information.
\end{enumerate}

Now consider a conceptual transformation of a \scheme'wcm' directive, \scheme|C[(wcm mark-expr body-expr)]|, as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    ((C[body-expr] TRUE) (let ((mark-value ((C[mark-expr] FALSE) marks))
                               (rest-marks (if flag (snd marks) marks)))
                           (cons mark-value rest-marks)))))
\end{schemedisplay}
with similar caveats as the previous case. This definition captures that
\begin{enumerate}
\item as in application, we expect \scheme'flag' to indicate tail position information and \scheme'marks' to provide a list of the current continuation marks,
\item we evaluate \scheme'mark-expr' with correct contextual information,
\item we discard the first continuation mark of the parent context if evaluation is occurring in tail position of a \scheme'wcm' directive,
\item and we evaluate \scheme'C[body-expr]' with the correct tail-position flag and current continuation marks.
\end{enumerate}

Finally, consider the conceptual transformation of a \scheme'ccm' directive, \scheme|C[(ccm)]|, as
\begin{schemedisplay}
(lambda (flag)
  (lambda (marks)
    marks))
\end{schemedisplay}
wherein we reap the fruits of simplicity from our laborious passing: this definition is gratifyingly direct.

@@

We can express the \scheme'let' construct in \lv\ with application. An expression such as
\begin{schemedisplay}
(let ((x_1 e_1)
      ...
      (x_n e_n))
  e)
\end{schemedisplay}
can be interpreted as
\begin{schemedisplay}
(...((lambda (x_1) ... (lambda (x_n)
           e)...) e_1) ...) e_n)
\end{schemedisplay}
which is the curried form of
\begin{schemedisplay}
((lambda (x_1 ... x_n)
           e) e_1 ... e_n)
\end{schemedisplay}
We use this fact to construct the transformation and apply $\beta$-rule to simplify terms statically.

@@

Abstracting terms has the effect of suspending evaluation. When an entire program is transformed, all evaluation is suspended, and awaits arguments representing contextual information. At the top level, the context is empty, so we pass the contextual information for the empty context: \scheme'FALSE' indicating evaluation is \emph{not} occurring in \scheme'wcm' tail position and \scheme'NIL', an empty list of marks.

@@

For this reason, we adjust $\mathcal{C}'$ so that
\begin{equation}
\Cp{p}=\app{\app{\C{p}}{\false}}{\C{\nil}}
\end{equation}
and recognize that a fully-realized transform must be made aware of this.

@@

@@

\begin{schemedefn}{\scheme|C[(wcm mark-expr body-expr)]|}
\begin{schemedisplay}
(lambda (kont)
  (lambda (flag)
    (lambda (marks)
      (((C[mark-expr]
         (lambda (mark-value) ((lambda (k_1) 
                                 (k_1 (((marks (lambda (x) x)) (lambda (z) z)) (lambda (z) z))))
                               (lambda (t) 
                                 ((lambda (k_2) (k_2 ((f C[(SND t)]) t)))
                                   (lambda (rest) 
                                     (((C[body-expr] kont) TRUE) C[(CONS mark-value rest)])))))))
        FALSE)
       marks))))
\end{schemedisplay}
\end{schemedefn}
